---
title: "PA1_Reproducible_Research"
author: "Mark"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Code for reading in the dataset and/or processing the data

```{r activit, echo=TRUE}
library(readr)
activity <- read.csv("activity.csv")
head(activity)
tail(activity)
str(activity)
summary(activity)
```

## Step 2 - Histogram of the total number of steps taken each day
```{r plot, echo=TRUE}
library(dplyr)
library(ggplot2)
daily_sum <- activity %>% group_by(date) %>% summarise(total_steps = sum(steps), na.rm = TRUE)
ggplot(data = daily_sum, aes(x = total_steps)) +
  geom_histogram(binwidth = 1000, fill = "cornflowerblue", color = "white") +
  labs(x = "Total Steps per Day", y = "Frequency", title = "Histogram of Total Steps per Day")

```
## Step 3 - Mean and median number of steps taken each day
```{r mean, echo=TRUE}
activity %>% group_by(date) %>% summarise(mean_steps = mean(steps, na.rm = TRUE), median_steps = median(steps, na.rm = TRUE))

```
## Step 4 -  Time series plot of the average number of steps taken
```{r time_series plot, echo=TRUE}
mean_steps <- activity %>% group_by(date) %>% summarise(mean_steps = mean(steps, na.rm = TRUE))
ggplot(mean_steps, aes(x = date, y = mean_steps))+
    geom_bar(stat = "identity", fill = "blue") +
    labs(x = "Date", y = "Mean_Steps", title = "Time Series Mean Steps")

```
## Step 5 - The 5-minute interval that, on average, contains the maximum number of steps
- transform column interval to character type
- calculate the mean steps for each interval
- sort the steps in descending for interval
- select the first row in the data set as the maximum

```{r max interval}
activity$interval <- as.character(activity$interval)
mean_interval_steps <- activity %>% group_by(interval) %>% summarise(mean_steps = mean(steps, na.rm = TRUE))
sorted_data <- mean_interval_steps %>% arrange(desc(mean_steps))
sorted_data %>% slice(1)

```
## Step 6 -  Code to describe and show a strategy for imputing missing data
using sum to understand the number of NA in the column steps, mean() to get the % of NA
calculate the mean value by group of interval and acquire a new data set 
merge the activity data set with the merged_activity one by the column of interval 
replace the NA by the mean value for respective interval group
remove the mean_steps column 
transform the numeric type to integer for easier understanding
```{r}
sum(is.na(activity$steps))
mean(is.na(activity$steps))
activity$date <- as.character(activity$date)
act_mean <- activity %>% group_by(interval) %>% summarise(step_mean = mean(steps, na.rm = TRUE))
merged_activity <- merge(activity, act_mean, by = "interval")
merged_activity$steps[is.na(merged_activity$steps)] <- merged_activity$step_mean[is.na(merged_activity$steps)]
merged_activity$step_mean <- NULL
merged_activity$steps <- as.integer(merged_activity$steps)
summary(merged_activity)

```
## Step 7 - Histogram of the total number of steps taken each day after missing values are imputed
after removing the missing value, the distribution of today steps by date is a bit different 
the mean value is a higher than the original one
```{r imputed_NA_hist, echo=TRUE}
daily_sum_2 <- merged_activity %>% group_by(date) %>% summarise(total_steps = sum(steps), na.rm = TRUE)
ggplot(data = daily_sum_2, aes(x = total_steps)) +
  geom_histogram(binwidth = 1000, fill = "cornflowerblue", color = "white") +
  labs(x = "Total Steps per Day", y = "Frequency", title = "Histogram of Total Steps per Day")

```
## Step 8 - Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r weekday_weekend, echo=TRUE}
merged_activity$date <- as.Date(merged_activity$date)
merged_activity$day_type <- ifelse(weekdays(merged_activity$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
merged_activity$day_type <- factor(merged_activity$day_type, levels = c("weekday", "weekend"))
steps_day_type <-  aggregate(steps ~ interval + day_type, merged_activity, mean)
steps_day_type$interval <- as.integer(steps_day_type$interval)
ggplot(steps_day_type, aes(interval, steps, color = day_type)) + 
  geom_line() + ggtitle("Mean Daily Steps per Day Type") + 
  xlab("Interval") + 
  ylab("Number of Steps") +
  facet_wrap(.~day_type, ncol=1,nrow=2) 

```































